Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2020-12-21 journal paper/content/related_work.tex
pass through alex's edits

\xhdr{Exploration and competition.}
\citet{bergemann1997market,bergemann2000experimentation,keller2003price} study the interplay of exploration and competition for users when the competing firms experiment with \emph{prices}
\asedit{(whereas in our model they experiment with design alternatives).} 
All three papers consider strategies that respond to competition and analyze Markov-perfect equilibria (whereas we focus on the adoption of better bandit algorithms). We discuss them in more detail below.

In \citet{keller2003price}, the competition is entirely on prices, rather than product quality. This distinction leads to several important differences. First, the data externality -- that each customer brings a new data point to a firm if and only if he chooses this firm -- is absolutely crucial to our setting, and absent in theirs. Second, the inputs to agents' decision rule are prices, rather than the quality of the chosen alternatives; the former is directly controlled by the firms, whereas the latter is not known a priori. Third, the goal of exploration is to learn the agents' decision rule.

The earlier work by \citet{bergemann1997market,bergemann2000experimentation} studies competition on both prices and quality, and allows for the data externality mentioned above. Specifically, the ``entrant" firm offers a new product of unknown, exogenously determined, quality, which is ``explored" if and only if the entrant attracts customers. However, because the entrant can only control the prices but not the product itself, the nature of exploration is fundamentally different. First, the goal of exploration is to reveal information about a fixed product. Second, the entrant can only affect the \emph{quantity} of the said information, but not \emph{which} information is being revealed. Third, revealing more information is not necessarily better for the entrant (if the product's quality is actually low). Fourth, the social planner's exploration problem is very different from the firms', so that when one compares competition to the planner, one can only compare outcomes, but not the algorithms.\footnote{The planner faces a two-armed bandit problem, because the planner can directly choose a firm for each agent.}
% firm, because payments cancel out in social welfare).}

@@

The literature on learning-by-doing vs. competition \citep[\eg][]{fudenberg1983learning, dasgupta1988learning, cabral1994learning} 
\asmargincomment{Guy: pls cite Gilbert-Newberry, as per YK.}
studies firms that learn while competing against each other, so that a firm attracting more consumers reduces its production costs. Our model differs in several important respects. First, firms learn to improve product quality rather than to reduce production costs. Second, the firms' current actions have consequences (via reputation and/or data collected by the algorithm) that directly impact consumer choices in the future. Third, we endogenize the technology behind learning-by-doing by explicitly considering bandit algorithms.

@@

\xhdr{Choice models.}
Stochastic choice models similar to ours have been widely used throughout economics. First, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. They play a similar role in our model, side-stepping the dominance of the greedy algorithm.
%as we move from \HardMax to \HardMaxRandom.
Second, there a large literature on non-existence of equilibria due to small deviations, starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets.
\asdelete{\footnote{\cite{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom in this context.}}
This is superficially similar to how small deviations towards the greedy algorithm rule out equilibria under \HardMaxRandom.
%Second, the stochastic choice rules we consider have been widely utilized across economics. For instance,
Third, \SoftMaxRandom subsumes the logit choice rule, a standard behavioral model with strong empirical and microeconomic foundations
\citep[\eg][]{mosteller1951experimental, luce1959choice, mcfadden1974measurement, matvejka2015rational}.
Fourth, choice models similar to \SoftMaxRandom are used to explain  horizontal product differentiation \citep[\eg][]{Hotelling-29, Perloff-Salop-85}.
%\asdelete{While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}

>>>>

\xhdr{Exploration and competition.}
\citet{bergemann1997market,bergemann2000experimentation,keller2003price} study the interplay of exploration and competition for users when the competing firms experiment with \emph{prices}
(whereas in our model they experiment with design alternatives).
All three papers consider strategies that respond to competition and analyze Markov-perfect equilibria (whereas we focus on the adoption of better bandit algorithms). We discuss them in more detail below.

In \citet{keller2003price}, the competition is entirely on prices, rather than product quality. This distinction leads to several important differences. First, the data externality -- that each customer brings a new data point to a firm if and only if he chooses this firm -- is absolutely crucial to our setting, and absent in theirs. Second, the inputs to agents' decision rule are prices, rather than the quality of the chosen alternatives; the former is directly controlled by the firms, whereas the latter is not known a priori. Third, the goal of exploration is to learn the agents' decision rule, \gaedit{not the best product alternative}.

The earlier work by \citet{bergemann1997market,bergemann2000experimentation} studies competition on both prices and quality, and allows for the data externality mentioned above. Specifically, the ``entrant" firm offers a new product of unknown, exogenously determined, quality, which is ``explored" if and only if the entrant attracts customers. However, because the entrant can only control the prices but not the product itself, the nature of exploration is fundamentally different \gaedit{than in our setting}. First, the goal of exploration is to reveal information about a fixed product. Second, the entrant can only affect the \emph{quantity} of the said information, but not \emph{which} information is being revealed. Third, revealing more information is not necessarily better for the entrant (if the product's quality is actually low). Fourth, the social planner's exploration problem is very different from the firms', so that when one compares competition to the planner, one can only compare outcomes, but not the algorithms.\footnote{The planner faces a two-armed bandit problem, because the planner can directly choose a firm for each agent.}
% firm, because payments cancel out in social welfare).}

@@

The literature on learning-by-doing vs. competition \citep[\eg][]{gilbert1982preemptive, fudenberg1983learning, dasgupta1988learning, cabral1994learning} 
\asmargincomment{Guy: pls cite Gilbert-Newberry, as per YK.}
studies firms that learn while competing against each other, so that a firm attracting more consumers reduces its production costs. Our model differs in several important respects. First, firms learn to improve product quality rather than to reduce production costs. Second, the firms' current actions have consequences (via reputation and/or data collected by the algorithm) that directly impact consumer choices in the future. Third, we endogenize the technology behind learning-by-doing by explicitly considering bandit algorithms.

@@

\xhdr{Choice models.}
Stochastic choice models similar to ours have been widely used throughout economics. First, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. They play a similar role in our model, side-stepping the dominance of the greedy algorithm.
%as we move from \HardMax to \HardMaxRandom.
\gadelete{Second, there a large literature on non-existence of equilibria due to small deviations, starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets.
\asdelete{\footnote{\cite{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom in this context.}}
This is superficially similar to how small deviations towards the greedy algorithm rule out equilibria under \HardMaxRandom.}
%Second, the stochastic choice rules we consider have been widely utilized across economics. For instance,
Second, \SoftMaxRandom subsumes the logit choice rule, a standard behavioral model with strong empirical and microeconomic foundations
\citep[\eg][]{mosteller1951experimental, luce1959choice, mcfadden1974measurement, matvejka2015rational}.
Third, choice models similar to \SoftMaxRandom are used to explain  horizontal product differentiation \citep[\eg][]{Hotelling-29, Perloff-Salop-85}.
%\asdelete{While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}

### 2020-12-21 journal paper/content/sec-bg.tex
pass through alex's edits

\asedit{Below we discuss which algorithms are better than others, and what does it \emph{mean} for one bandit algorithm to be better than another. This is a rather subtle issue, because some algorithms may be better for some problem instances and/or time intervals, and worse for some others. In particular, ``better" algorithms are better in the long run, but could be worse initially.}

>>>>

Below we discuss which algorithms are better than others, and what does it \emph{mean} for one bandit algorithm to be better than another. This is a rather subtle issue, because some algorithms may be better for some problem instances and/or time intervals, and worse for some others. In particular, ``better" algorithms are better in the long run, but could be worse initially.

### 2020-12-21 journal paper/content/sec-intro.tex
pass through alex's edits

In this paper, we study the interplay between \exploration and \competition.%
\footnote{\Ie we add \competition to the standard exploration-exploitation tradeoff studied in multi-armed bandits.}
Platforms that engage in exploration typically need to compete against one another. Most importantly, platforms compete for users, who benefit them in two ways:
generating revenue and providing data for learning. This creates a tension:
%between \exploration and \competition.
while exploration may be essential for improving the service tomorrow, it may degrade the \asedit{service} quality \emph{today}, in which case some of the users can leave and there will be fewer users to learn from. This may create a ``data feedback loop" when the platform's performance further degrades relative to competitors who keep learning and improving from \emph{their} users, and so forth. Taken to the extreme, such dynamics may cause a ``death spiral" effect when the vast majority of customers eventually switch to competitors.

@@

%We investigate these questions with
\xhdr{Our model: competition game.} We consider a stylized duopoly model in which two firms (\emph{principals}) compete for users (\emph{agents}). Principals compete on quality rather than on prices, and engage in exploration in order to learn which actions lead to high quality products. A new agent arrives and chooses a principal according to some decision rule. The principal selects an action \asedit{which affects the quality of service provided to this agent} (\eg a list of web search results). The agent experiences \asedit{this action} and reports a reward. Each principal only observes its own users. Principals commit to their strategies in advance, so as to maximize their market share.

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a multi-armed bandit algorithm, 
\asedit{which dynamically adjusts to the observed rewards.}%
\footnote{\asedit{We do not distinguish between pure and mixed strategies, because bandit algorithms can be randomized, and a distribution over bandit algorithms is also a bandit algorithm.}}
\asedit{However, it is oblivious to all signals on competition (such as the market share or the competitor's choices or rewards), even when such signals are available.}
%\footnote{\gaedit{In particular, the algorithm does not respond to neither the competitors' rewards (which it does not observe) nor its current market share.}} 
This modeling choice reflects the reality of industrial applications, which follow a huge body of knowledge in machine learning; more on this in Section~\ref{sec:discussion}. 
\asedit{Due to similar practical considerations, we expect the actual strategy choice to reflect the competition game only in a crude, qualitative way. Hence,  basic outcomes under competition are worth studying \emph{per se}, not only as a stepping stone to equilibrium characterization.}
%Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical.

\asedit{To flesh out the meaning of \emph{better} exploration algorithms, as per the main question, we draw on the literature from machine learning which studies such algorithms. In particular, one considers algorithm's performance \emph{in isolation}: in a standalone exploration problem without competition. We focus on big, qualitative differences between algorithms that are well-established in the literature, see Appendix~\ref{app:bg} for self-contained background.
%We draw on the machine learning literature to compare bandit algorithms to one another in isolation, so that we can talk about \emph{better} bandit algorithms in a principled way.\footnote{\gaedit{This literature typically compares the performance of different algorithms in a stand-alone exploration problem according to their asymptotic regret, which can be interpreted as maximizing consumer welfare in our context (see Appendix~\ref{app:bg} for self-contained background). Thus, we can utilize this comparison measure to assess the quality of the algorithms adopted under competition.}}
%\footnote{Such comparisons are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. More on this in Appendix~\ref{app:bg}. Generally,  ``better" algorithms are better in the long run, but could be worse initially. }
One baseline is algorithms that do not purposefully explore, and instead strive to make myopically optimal decisions; we call them \emph{greedy algorithms}. In isolation, they are known to perform poorly for a wide variety of problem instances.}

@@

\xhdr{Theoretical results.}
%We endow agents with Bayesian rationality, a common modeling approach for a theoretical investigation.
We consider a basic Bayesian model (called the \emph{\TheoryModel}), where agents have a common Bayesian prior on reward distributions, know the principals' algorithms \asedit{and their own arrival times}, but do not \asedit{observe} the previous agents' choices or rewards. Each agent computes the Bayesian-expected reward for each principal, and uses these two numbers to decide which principal to choose.
%We focus on qualitative, asymptotic differences in the algorithms' performance.
Our results depend crucially on agents' decision rule:
%\begin{itemize}

@@

\textbf{(iii)} We further soften the decision rule so that the selection probabilities vary smoothly in terms of principals' Bayesian-expected rewards;
%as a function of the difference between  principals' Bayesian-expected rewards;
We call it \SoftMaxRandom, a more realistic middle ground between \HardMax agents and random agents.%
\footnote{Alternatively, one can obtain a \SoftMaxRandom decision rule using a mixture of more ``basic" agent types that follow \HardMax unless the principal's Bayesian-expected rewards are too close to each other.}
In the most technical result of the paper, we find that a \asedit{sufficiently} better algorithm prevails under much weaker assumptions.
% on what constitutes a better algorithm.
%This is the most technical result of the paper.

@@

\xhdr{Numerical simulations.}
We consider a basic frequentist model (called the \emph{\ExptsModel}), where agents observe signals about the principals' past performance and 
\asedit{make their decisions naively,}
%base their decisions on these signals alone, 
without invoking any prior knowledge or beliefs. The performance signals are abstracted and aggregated as a scalar \emph{reputation score} for each principal, modeled as a sliding window average of its rewards. Thus, agents' decision rule depends only on the two reputation scores.
% We refer to this variant as the \emph{\ExptsModel}.%
%\footnote{In comparison, the theoretical results focus on another extreme, with Bayesian rationality and no performance signals.}
We refine and expand the theoretical results in several ways:

@@

\textbf{(iii)} 
\asedit{However, the greedy algorithm is sometimes \emph{not} the best strategy under high levels of competition.\footnote{In our theoretical results on \HardMax, the greedy algorithm is always the best strategy, mainly because it is aware of the Bayesian prior (whereas in the simulations  the prior is not available).}
We revisit algorithms' performance without competition, and we find that the most natural performance measure does not explain this phenomenon. We suggest a new, more nuanced performance measure that does.}

@@

\xhdr{Interpretation: network effects of data.}
Our model speaks to policy discussions on regulating data-intensive digital platforms \citep{furman2019unlocking, scott2019committee}, and particularly to the ongoing debate on the role of data in the digital economy. One fundamental question in this debate is whether data can serve a similar role as traditional ``network effects", creating scenarios when only one firm can function in the market \citep{Rysman09, jullien2019economics}.
%whereby, when these effects are present, in many cases only one firm can function in the market, leading to competition \emph{for} the market being more important than competition \emph{in} the market \citep{Rysman09, jullien2019economics}.
The death spiral/amplification effects mentioned above have a similar flavor: a relatively small \asedit{performance loss due to} exploration (resp., data advantage)  gets amplified under competition and causes the firm to be starved of users (resp., take over most of the market).
%\gaedit{We further find that a small data advantage for one firm gets amplified under competition and leads to that firm taking the entire market, showing that data can provide a similar incumbency advantage as those provided by traditional network effects and can serve as a barrier to entry in online markets.}
However, a distinctive feature of our approach is that we explicitly model the learning problem of the firms and consider them deploying algorithms for solving this problem.  Thus, we do not explicitly model the network effects, but they arise endogenously from our setup.

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of observable data. Approaches such as \citet{lambrecht2015can, bajari2018impact, varian2018artificial} argue that the diminishing returns to scale and scope of data in isolation mitigate such data feedback loops,
%as non-existent
but ignore the differences induced by learning in isolation versus under competition. Explicitly modeling the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified by the increased \emph{quality} of data gathered by better learning algorithms, not just the quantity thereof. In particular, we find that incumbency is good for innovation \asedit{and welfare, \emph{and}} % but also
creates a barrier to entry, 
\asedit{all} % precisely
due to data feedback loops.

\asedit{Interestingly, we find that data feedback loops can be bad for innovation (as in ``death spiral"), or good for innovation (as in ``amplification effect"), depending on the level of competition.}

>>>>

In this paper, we study the interplay between \exploration and \competition.%
\footnote{\Ie we add \competition to the standard exploration-exploitation tradeoff studied in multi-armed bandits.}
Platforms that engage in exploration typically need to compete against one another. Most importantly, platforms compete for users, who benefit them in two ways:
generating revenue and providing data for learning. This creates a tension:
%between \exploration and \competition.
while exploration may be essential for improving the service tomorrow, it may degrade the service quality \emph{today}, in which case some of the users can leave and there will be fewer users to learn from. This may create a ``data feedback loop" when the platform's performance further degrades relative to competitors who keep learning and improving from \emph{their} users, and so forth. Taken to the extreme, such dynamics may cause a ``death spiral" effect when the vast majority of customers eventually switch to competitors.

@@

%We investigate these questions with
\xhdr{Our model: competition game.} We consider a stylized duopoly model in which two firms (\emph{principals}) compete for users (\emph{agents}). Principals compete on quality rather than on prices, and engage in exploration in order to learn which actions lead to high quality products. A new agent arrives and chooses a principal according to some decision rule \gaedit{which depends on the principal's exploration strategy}. The principal selects an action which affects the quality of service provided to this agent (\eg a list of web search results). The agent experiences this action \gaedit{and the resulting utility from this action is observed by the principal.} \gadelete{and reports a reward.} Each principal only observes its own users. Principals commit to their strategies in advance, so as to maximize their market share.

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a multi-armed bandit algorithm, 
which dynamically adjusts to the observed rewards.%
\footnote{We do not distinguish between pure and mixed strategies, because bandit algorithms can be randomized, and a distribution over bandit algorithms is also a bandit algorithm.}
However, it is oblivious to all signals on competition (such as the market share or the competitor's choices or rewards), even when such signals are available.
%\footnote{\gaedit{In particular, the algorithm does not respond to neither the competitors' rewards (which it does not observe) nor its current market share.}} 
This modeling choice reflects the reality of industrial applications, which follow a huge body of knowledge in machine learning; more on this in Section~\ref{sec:discussion}. 
Due to similar practical considerations, we expect the actual strategy choice to reflect the competition game only in a crude, qualitative way. Hence,  basic outcomes under competition are worth studying \emph{per se}, not only as a stepping stone to equilibrium characterization.
%Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical.

To flesh out the meaning of \emph{better} exploration algorithms, as per the main question, we draw on the literature from machine learning which studies such algorithms. In particular, one considers algorithm's performance \emph{in isolation}: in a standalone exploration problem without competition. We focus on big, qualitative differences between algorithms that are well-established in the literature, see Appendix~\ref{app:bg} for self-contained background.
%We draw on the machine learning literature to compare bandit algorithms to one another in isolation, so that we can talk about \emph{better} bandit algorithms in a principled way.\footnote{\gaedit{This literature typically compares the performance of different algorithms in a stand-alone exploration problem according to their asymptotic regret, which can be interpreted as maximizing consumer welfare in our context (see Appendix~\ref{app:bg} for self-contained background). Thus, we can utilize this comparison measure to assess the quality of the algorithms adopted under competition.}}
%\footnote{Such comparisons are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. More on this in Appendix~\ref{app:bg}. Generally,  ``better" algorithms are better in the long run, but could be worse initially. }
One baseline is algorithms that do not purposefully explore, and instead strive to make myopically optimal decisions; we call them \emph{greedy algorithms}. In isolation, they are known to perform poorly for a wide variety of problem instances.

@@

\xhdr{Theoretical results.}
%We endow agents with Bayesian rationality, a common modeling approach for a theoretical investigation.
We consider a basic Bayesian model (called the \emph{\TheoryModel}), where agents have a common Bayesian prior on reward distributions, know the principals' algorithms and their own arrival times, but do not observe the previous agents' choices or rewards. Each agent computes the Bayesian-expected reward for each principal, and uses these two numbers to decide which principal to choose.
%We focus on qualitative, asymptotic differences in the algorithms' performance.
Our results depend crucially on agents' decision rule:
%\begin{itemize}

@@

\textbf{(iii)} We further soften the decision rule so that the selection probabilities vary smoothly in terms of principals' Bayesian-expected rewards;
%as a function of the difference between  principals' Bayesian-expected rewards;
We call it \SoftMaxRandom, a more realistic middle ground between \HardMax agents and random agents.%
\footnote{Alternatively, one can obtain a \SoftMaxRandom decision rule using a mixture of more ``basic" agent types that follow \HardMax unless the principal's Bayesian-expected rewards are too close to each other.}
In the most technical result of the paper, we find that a sufficiently better algorithm prevails under much weaker assumptions.
% on what constitutes a better algorithm.
%This is the most technical result of the paper.

@@

\xhdr{Numerical simulations.}
We consider a basic frequentist model (called the \emph{\ExptsModel}), where agents observe signals about the principals' past performance and 
make their decisions naively,
%base their decisions on these signals alone, 
without invoking any prior knowledge or beliefs. The performance signals are abstracted and aggregated as a scalar \emph{reputation score} for each principal, modeled as a sliding window average of its rewards. Thus, agents' decision rule depends only on the two reputation scores.
% We refer to this variant as the \emph{\ExptsModel}.%
%\footnote{In comparison, the theoretical results focus on another extreme, with Bayesian rationality and no performance signals.}
We refine and expand the theoretical results in several ways:

@@

\textbf{(iii)} 
However, the greedy algorithm is sometimes \emph{not} the best strategy under high levels of competition.\footnote{In our theoretical results on \HardMax, the greedy algorithm is always the best strategy, mainly because it is aware of the Bayesian prior (whereas in the simulations  the prior is not available).}
We revisit algorithms' performance \gaedit{in a standalone bandit problem, \ie without competition, and} we find that the most natural performance measure does not explain this phenomenon. We suggest a new, more nuanced performance measure that does.

@@

\xhdr{Interpretation: network effects of data.}
Our model speaks to policy discussions on regulating data-intensive digital platforms \citep{furman2019unlocking, scott2019committee}, and particularly to the ongoing debate on the role of data in the digital economy. One fundamental question in this debate is whether data can serve a similar role as traditional ``network effects", creating scenarios when only one firm can function in the market \citep{Rysman09, jullien2019economics}.
%whereby, when these effects are present, in many cases only one firm can function in the market, leading to competition \emph{for} the market being more important than competition \emph{in} the market \citep{Rysman09, jullien2019economics}.
The death spiral/amplification effects mentioned above have a similar flavor: a relatively small performance loss due to exploration (resp., data advantage)  gets amplified under competition and causes the firm to be starved of users (resp., take over most of the market).
%\gaedit{We further find that a small data advantage for one firm gets amplified under competition and leads to that firm taking the entire market, showing that data can provide a similar incumbency advantage as those provided by traditional network effects and can serve as a barrier to entry in online markets.}
However, a distinctive feature of our approach is that we explicitly model the learning problem of the firms and consider them deploying algorithms for solving this problem.  Thus, we do not explicitly model the network effects, but they arise endogenously from our setup.

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of observable data. Approaches such as \citet{lambrecht2015can, bajari2018impact, varian2018artificial} argue that the diminishing returns to scale and scope of data in isolation mitigate such data feedback loops,
%as non-existent
but ignore the differences induced by learning in isolation versus under competition. Explicitly modeling the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified by the increased \emph{quality} of data gathered by better learning algorithms, not just the quantity thereof. In particular, we find that incumbency is good for innovation and welfare, \emph{and} % but also
creates a barrier to entry, 
all % precisely
due to data feedback loops.

Interestingly, we find that data feedback loops can be bad for innovation (as in ``death spiral"), or good for innovation (as in ``amplification effect"), depending on the level of competition.

### 2020-12-21 journal paper/content/sec-model.tex
pass through alex's edits

\xhdr{Principals and agents.} There are two principals and $T$ agents. We denote them, resp., principal $i\in \{1,2\}$ and agent $t\in [T]$,
\asedit{where $[T] := \{1,2\LDOTS T\}$}.
The game proceeds in $T$ rounds.

In each round $t\in [T]$, the following  interaction takes place. Agent $t$ arrives and chooses a principal $i_t\in \{1,2\}$. The principal chooses action $a_t\in A$, where $A$ is a fixed set of actions.%
\footnote{We use `action' and `arm' interchangeably, as common in the literature on multi-armed bandits.}
The agent experiences this action and receives an associated reward $r_t\in \{ 0,1\}$, which is then observed by the principal. We posit \emph{stochastic rewards}: whenever a given action $a\in A$ is chosen, the reward is an independent draw from Bernoulli distribution with mean $\mu_a$. \asedit{In particular, the mean rewards $\mu_a$, as well as the action set $A$, are the same for both principals and all rounds.} The mean rewards are initially not known to anybody. The principals are completely unaware of the rounds when the opponent is chosen.

@@

Thus, each principal follows the protocol of \emph{multi-armed bandits} (henceforth, \emph{MAB}). Specifically, in each round when it is chosen, it chooses an arm from $A$ and observes a reward for this action. \asedit{It does not observe anything else, particularly rewards for other actions.}

Each principal $i$ commits to an MAB algorithm \alg[i] before round $1$, and uses this algorithm throughout the game. The algorithm proceeds in time-steps:
%\footnote{These time-steps will sometimes be referred to as \emph{local steps/rounds}, so as to distinguish them from ``global rounds" defined before. We will omit the global vs. local distinction when clear from the context.}
each time it is called, it outputs an arm from $A$, and inputs a reward for this action. The algorithm is called only in game rounds when principal $i$ is chosen.
\asedit{When the distinction between algorithm's time-steps and game rounds is unclear from the context, we will refer to them as, resp., \emph{local steps/rounds} and \emph{global rounds}.}

@@

\xhdr{Agents' response.} Each agent $t$ forms reward estimates $\est_i(t)\in [0,1]$ for each principal $i$. \asedit{(What these estimates are, and how much the agents know in order to form them, depends on the Bayesian vs. frequentist model variant.)
The reward estimates determine the choice of the principal.} Specifically, agent $t$ chooses principal $1$ with probability
\begin{align}\label{eq:model-f}
p_t = \respF\rbr{ \est_1(t) - \est_2(t) },
\end{align}
where $\respF:[-1,1]\to [0,1]$ is the \emph{response function}, same for all agents. We assume that $\respF$ is monotonically non-decreasing, is larger than $\nicefrac12$ on the interval $(0,1]$, and smaller than $\nicefrac12$ on the interval $[-1,0)$. We consider three variants for $\respF$, depicted in  \reffig{fig:response-functions}:

@@

\HardMaxRandom and \SoftMaxRandom can be seen as agent types that, resp., give some chance to the other principal and avoid sharp transitions in their probabilistic response. 
\asmargincomment{I don't get YK's beef with this.}
Alternatively, they can be realized as a \asedit{distribution} over more ``basic" agent types. Indeed, \HardMaxRandom \asedit{distribution} is a mixture of \HardMax and ``random agents" (which choose a principal uniformly at random). The latter can be interpreted as consumers oblivious to principals' reputation, due to the lack of awareness or interest. One can obtain a \SoftMaxRandom response function using agent types that choose a principal $i$ with a largest reward estimate $\est_i$, unless $|\est_1-\est_2|$ is upper-bounded by some parameter $\theta$, in which case they choose uniformly. Then, we obtain \SoftMaxRandom as a mixture of random agents and these ``$\theta$-\HardMax'' agents, for a suitable distribution over $\theta$.

@@

\xhdr{Bayesian vs. frequentist.}
We consider two model variants, Bayesian and frequentist \asedit{(which we use, resp., for theoretical results and numerical simulations)}. The main difference between the two concerns the agents' reward estimates $\est_i(t)$.

@@

\xhdr{Competition game.}
Some of our results explicitly study the game between the two principals, termed the \emph{competition game}. We model it as a simultaneous-move game: before the first agent arrives, each principal commits to an MAB algorithm. \asedit{(Since MAB algorithms can be randomized, we make no distinction between pure and mixed strategies.)}
 Principal's utility is defined as the market share, \ie the number of agents that chose this principal. Principals are risk-neutral and aim to maximize their expected utility.

@@

\asedit{Now that we laid out our model, let us discuss several aspects thereof that are unclear from the bare definitions. In particular, we flesh out several points made in the Introduction.}

>>>>

\xhdr{Principals and agents.} There are two principals and $T$ agents. We denote them, resp., principal $i\in \{1,2\}$ and agent $t\in [T]$,
where $[T] := \{1,2\LDOTS T\}$.
The game proceeds in $T$ rounds.

In each round $t\in [T]$, the following  interaction takes place. Agent $t$ arrives and chooses a principal $i_t\in \{1,2\}$. The principal chooses action $a_t\in A$, where $A$ is a fixed set of actions.%
\footnote{We use `action' and `arm' interchangeably, as common in the literature on multi-armed bandits.}
The agent experiences this action and receives an associated reward $r_t\in \{ 0,1\}$, which is then observed by the principal. We posit \emph{stochastic rewards}: whenever a given action $a\in A$ is chosen, the reward is an independent draw from Bernoulli distribution with mean $\mu_a$. In particular, the mean rewards $\mu_a$, as well as the action set $A$, are the same for both principals and all rounds. The mean rewards are initially not known to anybody. The principals are completely unaware of the rounds when the opponent is chosen.

@@

Thus, each principal follows the protocol of \emph{multi-armed bandits} (henceforth, \emph{MAB}). Specifically, in each round when it is chosen, it chooses an arm from $A$ and observes a reward for this action. It does not observe anything else, particularly rewards for other actions.

Each principal $i$ commits to an MAB algorithm \alg[i] before round $1$, and uses this algorithm throughout the game. The algorithm proceeds in time-steps:
%\footnote{These time-steps will sometimes be referred to as \emph{local steps/rounds}, so as to distinguish them from ``global rounds" defined before. We will omit the global vs. local distinction when clear from the context.}
each time it is called, it outputs an arm from $A$, and inputs a reward for this action. The algorithm is called only in game rounds when principal $i$ is chosen.
When the distinction between algorithm's time-steps and game rounds is unclear from the context, we will refer to them as, resp., \emph{local steps/rounds} and \emph{global rounds}.

@@

\xhdr{Agents' response.} Each agent $t$ forms reward estimates $\est_i(t)\in [0,1]$ for each principal $i$. (What these estimates are, and how much the agents know in order to form them, depends on the Bayesian vs. frequentist model variant.)
The reward estimates determine the choice of the principal. Specifically, agent $t$ chooses principal $1$ with probability
\begin{align}\label{eq:model-f}
p_t = \respF\rbr{ \est_1(t) - \est_2(t) },
\end{align}
where $\respF:[-1,1]\to [0,1]$ is the \emph{response function}, same for all agents. We assume that $\respF$ is monotonically non-decreasing, is larger than $\nicefrac12$ on the interval $(0,1]$, and smaller than $\nicefrac12$ on the interval $[-1,0)$. We consider three variants for $\respF$, depicted in  \reffig{fig:response-functions}:

@@

\HardMaxRandom and \SoftMaxRandom can be seen as agent types that\gadelete{, resp., give some chance to the other principal and avoid sharp transitions in their probabilistic response.} \gaedit{make mistakes due to being oblivious to the principals' reputation, due to lack of awareness or interest. \HardMaxRandom models users as either being aware of the reputation or not, whereas \SoftMaxRandom models user mistakes as following a smooth density.}
\asmargincomment{I don't get YK's beef with this.}
Alternatively, they can be realized as a distribution over more ``basic" agent types. Indeed, \gaedit{the} \HardMaxRandom distribution is a mixture of \HardMax and ``random agents" (which choose a principal uniformly at random). \gadelete{The latter can be interpreted as consumers oblivious to principals' reputation, due to the lack of awareness or interest.} One can obtain a \SoftMaxRandom response function using agent types that choose a principal $i$ with a largest reward estimate $\est_i$, unless $|\est_1-\est_2|$ is upper-bounded by some parameter $\theta$, in which case they choose uniformly. Then, we obtain \SoftMaxRandom as a mixture of random agents and these ``$\theta$-\HardMax'' agents, for a suitable distribution over $\theta$.

@@

\xhdr{Bayesian vs. frequentist.}
We consider two model variants, Bayesian and frequentist (which we use, resp., for theoretical results and numerical simulations). The main difference between the two concerns the agents' reward estimates $\est_i(t)$.

@@

\xhdr{Competition game.}
Some of our results explicitly study the game between the two principals, termed the \emph{competition game}. We model it as a simultaneous-move game: before the first agent arrives, each principal commits to an MAB algorithm. \gaedit{Principals are risk-neutral and their utility is defined as their market share, \ie the number of agents that chose this principal, and thus they aim to select the algorithm that maximizes their expected utility. (Since MAB algorithms can be randomized, we make no distinction between pure and mixed strategies.)}
\gadelete{Principal's utility is defined as the market share, \ie the number of agents that chose this principal. Principals are risk-neutral and aim to maximize their expected utility.}

@@

Now that we laid out our model, let us discuss several aspects thereof that are unclear from the bare definitions. In particular, we flesh out several points made in the Introduction.
